A comparison between the latest Prophet forecasting model and the ARIMA model: An evaluation of BTC/ZAR predictability

### 2017-10-27 Draft/Write-up/Draft.tex
Fixed about 80% of the mess. Still fixing the rest

\begin{center}
{\Huge \bf How does the latest Prophet forecasting model compare to the ARIMA
model? Evaluation of BTC/ZAR predictability through a Mincer-Zarnowitz
approach}\\
\today\\
Fiona Ganie (GNXFIO001)\\
{\tt GNXFIO001@myuct.ac.za}
\end{center}

\begin{abstract}
The Autoregressive Integrated Moving Average (ARIMA) model is one of the
most widely used time series models that has gained popularity in the
exchange rate market due to its ease of implementation and tractability.
With the evolution of computational power, soft computing techniques
have since been used in exchange rate markets. The ARIMA model has been
used as the standard benchmark model against which these complex methods
have been compared. Although they have produced significantly better
results than the ARIMA model, these models lack interpretability and
building them is a challenging task. Prophet is a sophisticated model
that differs to traditional time series models in that it can be
utilised by non-experts who have little knowledge about the statistical
intricacies involved in the model, however have domain knowledge about
the data generating process. Prophet is also robust to outliers and can
handle missing values in the time series without the need for
interpolation. This paper compares the out-of-sample forecasts of
Prophet and ARIMA over varying forecast horizons, using a
Mincer-Zarnowitz approach for forecast evaluation. A complete dataset of
the prices of Bitcoin/Rand as well as a dataset which has missing prices
and outliers present is used in the analysis. It is found that the ARIMA
model produces the most efficient and unbiased forecasts over all
forecast horizons, and that the ARIMA model is more robust when faced
with missing values and outliers present in the data.
\noindent
Keywords: Prophet, ARIMA, Mincer-Zarnowitz,   Box-Jenkins, Bitcoin.
\end{abstract}
\end{titlepage}

@@

The Autoregressive Integrated Moving Average (ARIMA) model is one of the
most widely used time series models that has attracted attention in
financial market forecasting (Khashei, Bijari, and Ardali 2009).
Although the Random Walk model has typically been applied to foreign
exchange markets and has produced superior results, some researchers
have contended that foreign exchange markets are not efficient and
believe that future prices depend on current and past events
(Abu-Mostafa and Atiya 1996). This has led to the application of the
ARIMA model to exchange rate problems, where it has since gained
popularity due to its ease of implementation and tractability. The ARIMA
model is easy to use and yields good forecasts over short forecast
horizons, however the linearity of the ARIMA model fails to adequately
capture the non-linearity inherent in exchange rate data (Zhang and Hu
1998).

With the evolution of computational power, non-linear, soft computing
techniques have been proposed as a solution. The ARIMA model has been
used as the standard benchmark model against which these more complex
methods have been compared. Although they have produced significantly
better results than the ARIMA model, these models lack interpretability
and building them is a challenging task. The ARIMA model is tractable
and less computationally expensive. It has been used as the building
blocks for more advanced models and has provided the inspiration for
hybrid versions of the model which have been used in exchange rate
forecasting.

In December 2016, Facebook open-sourced their forecasting model Prophet.
Prophet differs to traditional time series model such as the ARIMA, in
that it can produce high quality forecasts in a straightforward way.
Prophet is a sophisticated model that provides informative results,
however is configurable and easy to use. It can be utilised by
non-experts who have little knowledge about the statistical intricacies
involved in the model, however have domain knowledge about the data
generating process. This knowledge can be easily incorporated into the
model through its intuitively adjustable parameters (Taylor and Letham
2017). Furthermore, Prophet is robust to outliers and can handle missing
values in the time series without the need for interpolation (Taylor and
Letham 2017). This allows analysts without knowledge on how to
pre-process the data, to utilise the model without sacrificing
predictive accuracy. If Prophet produces forecasts that are as good as
the ARIMA model, it can be compared to more complex forecasting methods
that are less tractable and flexible.

This paper broadly aims to compare the forecasts produced by the
traditional ARIMA model and Prophet through an evaluation of the
exchange rate between Bitcoin and the Rand (BTC/ZAR), using a
Mincer-Zarnowitz approach of measuring forecast accuracy. More
specifically, this paper aims to:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Compare the out-of-sample forecasts of ARIMA and Prophet over varying
  forecast horizons.
\item
  Compare the out-of-sample forecast performance with missing values and
  outliers present in the data.
\end{enumerate}

Section \ref{Background and Theory} will present the key findings from
the major works in which comparisons have been made to the ARIMA model
in exchange rate forecasting, as well as the theory behind the ARIMA and
Prophet model. Section \ref{Data} will provide a brief explanation and
motivation for the dataset used in this paper. Section \ref{Methodology}
will describe the methodology used in selecting the ARIMA and Prophet
models and the Mincer-Zarnowitz approach of measuring forecast accuracy.
The results from applying the ARIMA and Prophet model to the dataset is
presented in Section \ref{Results}. Finally, Section
\textbackslash{}ref\{Discussion and Conclusions discusses the results in
a statistical context.

\section{\texorpdfstring{Background and Theory
\label{Background and Theory}}{Background and Theory }}\label{background-and-theory}

@@

ARIMA models became highly popular since its introduction by Box and
Jenkins, when it was shown that they could outperform complex
econometric models in a variety of situations (Hibon and Makridakis
1997). The ARIMA model expresses the process \{\(y_t\)\} as a function
of the weighted average of past values of the process and lagged values
of the residuals. The weighted average of the past p values of the
process represents an autoregressive (AR) process of order p.~It feeds
back past values of the process into the current value, inducing
correlation between all lags of the process. The weighted average of the
q lagged residuals represents a moving average (MA) process of order q.
The purpose of mixing the MA process with the AR process is to reduce
the large number of past values required by AR processes and to control
for the autocorrelation which it creates between lagged values of the
process. The combination of the AR(p) and MA(q) process results in a
more parsimonious model, and forms a stationary autoregressive moving
average (ARMA(p,q)) process defined as:

@@

The ARIMA(p,d,q) model generalises the ARMA model in that it includes
both stationary and non-stationary processes. The parameter d is the
degree of differencing required to render the process stationary. If d
is equal to zero the process is stationary and equivalent to an ARMA
model, and if d is strictly positive the process requires differencing
to make it stationary. The ARIMA model can be defined succinctly using
the backward shift operator B, which shifts the process back by one unit
of time, and is defined as \(By_t = y_{t-1}\). The ARIMA model has the
form:

@@

ARIMA models have commonly been used in financial forecasting and are
popular for observing stock prices and exchange rates due to its power
and statistical properties (C.-S. Lin, Chiu, and Lin 2012). They have
frequently been used as a benchmark to compare new forecasting
techniques that have emerged over time. Kamruzzaman and Sarker (2003)
applied an ARIMA model to forecast the exchange rate between the
Australian dollar and six other currencies, and used the forecast errors
as well as the accuracy of the direction of the forecasts to evaluate
the performance of three Neural Network models. In a similar study,
Khashei, Bijari, and Ardali (2012) compared the predictive capability of
their proposed hybrid model consisting of an ARIMA and a Probabilistic
Neural Network (PNN) to the traditional ARIMA model, and justified the
use of the ARIMA as a benchmark by claiming that it is the most
important linear model. Their comparison was made by investigating the
forecasts between the British pound and US dollar over varying forecast
horizons. Not only has the ARIMA model been useful as a benchmark, but
it has also yielded satisfactory results when predicting exchange rates.
When Nwankwo (2014) forecasted the rate between the Nairo and dollar,
diagnostic testing revealed that the ARIMA(1,0,0) model was the best fit
for the data based on Akaike's Information Criterion (AIC).

The ARIMA model is attractive as it is tractable and produces good
short-term forecasts when more than 100 observations are used (Tseng et
al. 2001). Although the ARIMA model has the advantage of ease of
implementation and flexibility, it fails to capture the non-linearity
and volatility present in exchange rate data. Over time, the ARIMA model
has evolved to cater for a wider variety of data and to compensate for
some of its shortcomings. The most popular versions of the ARIMA model
that has been implemented in exchange rate forecasting is the Seasonal
ARIMA (SARIMA) and Fractional ARIMA model (FARIMA).

The SARIMA model was introduced to capture the periodic behaviour of
data and extends the ARIMA class by including a term for seasonal
differencing. Etuk, Wokoma, and Moffat (2013) modelled the Naira/CFA
Franc exchange rate which exhibited monthly seasonality using an
additive SARIMA model, to demonstrate that it can be a useful fit for
exchange rate data which displays seasonality. Their results showed that
the SARIMA model adequately described the variation in the exchange rate
series. The ARFIMA model generalises the ARIMA model in that the degree
required to make the data stationary can assume any real value, and is
no longer restricted to the integer domain. The ARFIMA model has the
ability to capture the dependence between observations that are widely
spread apart in time (Cheung 1993). This makes the model more
parsimonious since it can capture long memory in data as well as short
term dynamics (Cheung 1993). Cheung (1993) fitted the ARFIMA model to
examine five exchange rates, and found that there was strong evidence of
long memory in the exchange rate time series.

Generalised autoregressive conditional heteroskedasticity (GARCH) models
were later developed in an attempt to capture the volatility in
financial markets (Anastasakis and Mort 2009). Hsieh (1989) applied a
GARCH model to investigate five exchange rates and his results showed
that although the GARCH model outperformed the random walk, some
non-linear information still remained in the residuals.

Over time, financial forecasting methods have moved away from linear
models like ARIMA and GARCH, to soft computing techniques. These more
complex techniques are non-linear and can fit complex time series more
easily (Castillo and Melin 2002). Unlike the ARIMA model, soft computing
techniques do not impose structural assumptions on the model apriori
(Castillo and Melin 2002). Some of the most commonly used artificial
intelligence methods used to forecast exchange rate data are Neural
Networks and Fuzzy Logistic Systems.

Artificial Neural Networks (ANNs) have had many successful applications
in forecasting exchange rates and are more advantageous than other
non-linear forecasting methods. They are data driven, can adapt to
non-stationary environments and can approximate any continuous function
(Khashei and Bijari 2011). In a study done by Fahimifard et al. (2009),
the ANN was found as an effective way to improve the forecasts of
exchange rates. Superior results were produced when compared to the
ARIMA and GARCH model using the root mean square error (RMSE), mean
square error (MSE) and mean absolute difference as a measure of
performance. Fuzzy Logistic Systems (FLSs) were initially developed to
solve problems involving linguistic terms, and have been successfully
used in financial forecasting (Khashei, Bijari, and Ardali 2009). Fuzzy
logic tries to imitate human reasoning and the decision-making process
and allows for finer rather than discrete decisions to be provided.
Santos, Costa, and Santos Coelho (2007) investigated how well FLSs and
ANNs perform compared to the traditional ARMA and GARCH model. He
examined the forecasts of Brazilian exchange rate returns by considering
different frequencies of the series and comparing their one step-ahead
forecasts. By analysing accuracy statistics, he found that FLSs and ANNs
achieved higher returns based on the forecasts they produced. Similar
results were found by Khashei, Bijari, and Ardali (2009) when he
analysed the predictive capabilities of FLSs, ANNs, the traditional
ARIMA model, and a Fuzzy ARIMA model.

Although ANNs have been broadly applied in financial forecasting, the
process of building them is a complex task and there is no consistent
method of design compared to the traditional Box-Jenkins ARIMA model.
Unlike ARIMA models, the performance of ANNs is sensitive to many
modelling factors such as the number of input nodes included and the
size of the training sample chosen (Zhang and Hu 1998). Like ANNs there
is no systematic approach to designing FLSs and they are only
understandable when simple. Although FLSs has the advantage over ARIMA
models that they can be applied to data with few observations available,
it gives acceptable rather than accurate results and are more suitable
for problems which do not require high accuracy.

\subsection{Hybrid ARIMA Models}\label{hybrid-arima-models}

Over time, many researchers began to think of ways in which to harness
the advantages of tractable linear models such as the ARIMA model, and
of more complex non-linear models. By combining the ARIMA model with
other forecasting methods, the advantages of both forecasting methods
are leveraged while simultaneously improving their limitations. Some of
the hybrid models which have commonly been used in exchange rate
forecasting are the Fuzzy ARIMA and ANN-ARIMA model.

The ARIMA model produces very accurate forecasts over short time
horizons however it has the limitation of requiring more than 100
observations of historical data to yield accurate results (Tseng et al.
2001). In a world that is constantly changing and with the rapid
advancement of technology, access to large amounts of historical data is
difficult to obtain. On the contrary, a fuzzy regression model requires
little historical data however produces wide prediction intervals if
extreme values are present in the data. The Fuzzy ARIMA model combines
the ARIMA and fuzzy regression model to exploit the advantages of both
models while simultaneously overcoming their limitations. Tseng et al.
(2001) proposed applying a Fuzzy ARIMA model to forecast the exchange
rate of Taiwan dollars to US dollars to demonstrate the model's
appropriateness and power. The Fuzzy ARIMA not only produced forecasts
that were superior to the ARIMA and fuzzy time series models but also
provided an upper and lower bound which can be used by decision makers
to determine the best and worst possible situations.

Although ARIMA models are powerful, they require non-stationary data to
be differenced and impose prior assumptions onto the distribution of the
data (Ince and Trafalis 2006). In contrast, machine learning techniques
such as ANNs do not impose any assumptions onto the data generating
process however, being data-driven, are sensitive to the number of input
nodes used. The ANN-ARIMA model draws on the strengths of both models
and overcomes these individual difficulties. Ince and Trafalis (2006)
created an ANN-ARIMA model by using the ARIMA model to determine the
number input nodes required by an ANN for three exchange rates. When the
ANN-ARIMA model was compared to the pure ARIMA model, the hybrid model
outperformed the ARIMA based on the MSE.

Although the traditional ARIMA models produce less superior forecasts
than its hybrid forms and other complex non-linear techniques, its
forecasts are still satisfactory. They are simple models that are easy
to implement and have a consistent method of model design and selection.
ARIMA models are also more robust and efficient than complex structural
models in relation to short-run forecasting. The fact that they have
been used as the foundation for more advanced models and have commonly
been used as a benchmark for comparison, justifies it as a good starting
point to compare it to Facebook's forecasting method, Prophet, that was
recently released.

The techniques that have been considered for exchange rate forecasting
thus far, require the analyst to have vocational knowledge about time
series. Prophet differs to traditional time series models in that it is
flexible and can be customised by a large number of non-experts who have
little knowledge about time series, however have domain knowledge about
the data generating process. Prophet allows for a large number of
forecasts to be produced across a variety of problems and consists of a
robust evaluation system that allows for a large number of forecasts be
evaluated and compared. This is Facebook's definition of forecasting at
scale. Prophet is similar to a Generalized Additive Model (GAM) - an
additive regression model that consists of non-linear and linear
regression functions applied to predictor variables (Taylor and Letham
2017). The decomposable model is of the form:

@@

Prophet, like the GAM, frames the forecasting problem as a curve fitting
exercise and uses backfitting to find the regression functions. This
allows for the model to be fitted quickly and missing values and large
outliers to be handled elegantly. The regression model also provides
model flexibility and allows the analyst to interactively change model
parameters (Taylor and Letham 2017).

The growth component of Prophet may be modelled as a linear or
non-linear function of time. Linear growth is modelled by a piecewise
constant function while non-linear growth is modelled similar to
population growths which use a logistic growth model (Taylor and Letham
2017). A time-varying upper limit may be specified for logistic growth,
at which point the forecasts will saturate. This carrying capacity
allows the analyst to incorporate their prior knowledge about the
maximum obtainable growth level such as the total market or population
size into the model. Prophet accounts for changes in the trajectory of
this trend by automatically detecting and selecting changepoints in the
data at which the growth rate is allowed to change. These changepoints
have a Laplace prior distribution placed on them and its scale parameter
may be used to adjust the flexibility of the trend and to choose how
aggressively the model should follow historical trend changes. Analysts
may also adjust the number of potential changepoints included or
manually specify their location. This allows non-experts with knowledge
about events that may affect the growth rate to use the parameter as a
knob to either increase or decrease the number of changepoints included
(Taylor and Letham 2017). It also allows for the analyst to add
changepoints which the automatic selection procedure may have missed or
remove changepoints when the model is overfitting historical trends
(Taylor and Letham 2017).

@@

Prophet's Bayesian approach to forecasting allows the analyst to
incorporate their expert knowledge into the model building process and
has produced significantly improved forecasts compared to the ARIMA
model. Taylor and Letham (2017) forecasted the number of events on
Facebook using Prophet. The time series was impacted by holidays, had
strong multi-period seasonality, and a piecewise trend. The forecasts
produced by Prophet were compared to common forecasting techniques such
as exponential smoothing, ARIMA, the seasonal naïve, and the naïve
model. While exponential smoothing and the seasonal naïve model were
quite robust, the ARIMA forecasts were fragile. No model besides Prophet
accounted for the dips around holidays and the upward trend of the time
series towards later observations.

@@

The data set used in this study consists of the daily closing prices of
BTC/ZAR over the period 24 January 2016 to 17 July 2017. This comprises
of a total of 541 trading days and was the chosen time period for
analysis due to constraints in obtaining data over a longer period. The
data was obtained from Bitcoincharts.

Bitcoin is of specific interest as it presents an interesting parallel
to traditional exchange rate markets. The cryptocurrency is built on a
decentralised system and as a result its value cannot be directly
influenced by a central authority (Fantazzini and Nigmatullin 2016). In
addition, the Bitcoin market has attracted attention worldwide and is
currently the leading cryptocurrency, with awareness and adoption of the
currency growing over time (Fantazzini and Nigmatullin 2016). Its
novelty makes it a highly volatile hence speculative market and provides
an opportunity for forecasting.

@@

In this paper, the out-of-sample forecasts of ARIMA and Prophet are
compared over varying forecast horizons. The data is first split into a
training set and test set. The training set starts on the 24th January
2016 and ends on the 16th January 2017, while the test set starts on the
17th January 2017 and ends on the 17th July 2017. The Box-Jenkins
methodology is then used to select the correct ARIMA model while
Prophet's semi-automatic procedure is used for model selection. The
selected ARIMA and Prophet model are then fitted to the training set and
rolling window forecasts are made 1 day, 30 days and 90 days ahead. This
allows us to examine the forecast horizon effect. The Mincer-Zarnowitz
test is then used to evaluate the forecasts produced by ARIMA and
Prophet against the test set. Since there is no consensus on which
accuracy statistic best measures the performance of forecasting
techniques, the most popular criteria, the Mean Squared Error (MSE) is
employed. The results obtained from the ARIMA model will be used as a
benchmark for comparison. This process is then repeated based on data
which has outliers and missing values present.

@@

Box and Jenkins proposed a set of guidelines that can be followed when
selecting ARIMA models. This systematic procedure to designing ARIMA
models has made them highly popular (Hibon and Makridakis 1997). It
consists of a four-stage iterative process in which:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The process is either transformed or differenced to de-trend and
  stabilise the variance of the data.
\item
  The autocorrelation and partial autocorrelation plots are used to
  determine the order of p and q.
\item
  The parameters of the model are then estimated.
\item
  A diagnostic check is performed to ensure that the residuals are a
  white noise process.
\end{enumerate}

If the residuals are not white noise, steps 2-4 are repeated until a
satisfactory model is identified. On the contrary, if the diagnostic
check shows that the residuals are random, the developed model will be
the final model used for forecasting.

>>>>

\begin{center}
{\Huge \bf A comparison between the latest Prophet forecasting model and the ARIMA
model: An evaluation of BTC/ZAR predictability}\\
\today\\
Fiona Ganie (GNXFIO001)\\
{\tt GNXFIO001@myuct.ac.za}
\end{center}

\begin{abstract}
The Autoregressive Integrated Moving Average (ARIMA) model is one of the
most widely used time series models that has gained popularity in the
exchange rate market due to its ease of implementation and tractability.
With the evolution of computational power, soft computing techniques
have since been used in exchange rate markets. The ARIMA model has been
used as the standard benchmark model against which these complex methods
have been compared. Although they have produced significantly better
results than the ARIMA model, these models lack interpretability and
building them is a challenging task. Prophet is a sophisticated model
that differs from traditional time series models in that it can be
utilised by non-experts who have little knowledge about the statistical
intricacies involved in the model, however have domain knowledge about
the data generating process. Prophet is also robust to outliers and can
handle missing values in the time series without the need for
interpolation. This paper compares the out-of-sample forecasts of
Prophet and ARIMA over varying forecast horizons, using a
Mincer-Zarnowitz approach for forecast evaluation. A complete dataset of
the prices of Bitcoin/Rand as well as a dataset which has missing prices
and outliers present is used in the analysis. It is found that the ARIMA
model produces the most efficient and unbiased forecasts over all
forecast horizons, and that the ARIMA model is more robust when faced
with missing values and outliers present in the data.
\noindent\\
Keywords: Prophet, ARIMA, Mincer-Zarnowitz,   Box-Jenkins, Bitcoin, Model Confidence Set
\end{abstract}
\end{titlepage}

@@

The Autoregressive Integrated Moving Average (ARIMA) model is one of the
most widely used time series models that have attracted attention in
financial market forecasting (Khashei, Bijari, and Ardali 2009).
Although the Random Walk model has typically been applied to foreign
exchange markets and has produced superior results, some researchers
have contended that foreign exchange markets are not efficient and
believe that future prices depend on current and past events
(Abu-Mostafa and Atiya 1996). This has led to the application of the
ARIMA model to exchange rate problems, where it has since gained
popularity due to its ease of implementation and tractability. The ARIMA
model is easy to use and yields good forecasts over short forecast
horizons, however the linearity of the ARIMA model fails to adequately
capture the non-linearity inherent in exchange rate data (Zhang and Hu
1998).

With the evolution of computational power, non-linear, soft computing
techniques have been proposed as a solution. The ARIMA model has been
used as the standard benchmark model against which these more complex
methods have been compared. Although they have produced significantly
better results than the ARIMA model, these models lack interpretability
and building them is a challenging task. The ARIMA model is tractable
and less computationally expensive. It has been used as the building
block for more advanced models and has provided the inspiration for
hybrid versions of the model which have been used in exchange rate
forecasting.

In December 2016, Facebook open-sourced their forecasting model Prophet.
Prophet differs from traditional time series model such as the ARIMA in
that it can produce high quality forecasts in a straightforward way.
Prophet is a sophisticated model that provides informative results,
however is configurable and easy to use. It can be utilised by
non-experts who have little knowledge about the statistical intricacies
involved in the model, however have domain knowledge about the data
generating process. This knowledge can easily be incorporated into the
model through its intuitively adjustable parameters (Taylor and Letham
2017). Furthermore, Prophet is robust to outliers and can handle missing
values in the time series without the need for interpolation (Taylor and
Letham 2017). This allows analysts without knowledge on how to
pre-process the data, to utilise the model without sacrificing
predictive accuracy. If Prophet produces forecasts that are as good as
the ARIMA model, it can be compared to more complex forecasting methods
that are less tractable and flexible.

This paper broadly aims to compare the forecasts produced by the
traditional ARIMA model and Prophet through an evaluation of the
exchange rate between Bitcoin and the Rand (BTC/ZAR), using the Model
Confidence Set and a Mincer-Zarnowitz approach of measuring forecast
accuracy. More specifically, this paper aims at:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Comparing the out-of-sample forecasts of ARIMA and Prophet over
  varying forecast horizons.
\item
  Comparing the out-of-sample forecast performance with missing values
  and outliers present in the data.
\end{enumerate}

Section \ref{Background} will present the key findings from the major
works in which comparisons have been made to the ARIMA model in exchange
rate forecasting. Section \ref{Data} will provide a brief explanation
and motivation for the dataset used in this paper. Section
\ref{Forecasting Methods} explains and compares the formulation of the
ARIMA and Prophet model. Section \ref{Methodology} will describe the
methodology used in selecting the ARIMA and Prophet models, the Model
Confidence Set and the Mincer-Zarnowitz approach of measuring forecast
accuracy. The results from applying the ARIMA and Prophet model to the
dataset is presented in Section \ref{Results}. Finally, section
\ref{Discussion and Conclusions} discusses the results and its financial
implications in a statistical context.

\section{\texorpdfstring{Background
\label{Background}}{Background }}\label{background}

@@

ARIMA models have commonly been used in financial forecasting and are
popular for observing stock prices and exchange rates due to its power
and statistical properties (C.-S. Lin, Chiu, and Lin 2012). The ARIMA
model is attractive as it is tractable and produces good short-term
forecasts when more than 100 observations are used (Tseng et al. 2001).
They have frequently been used as a benchmark to compare new forecasting
techniques that have emerged over time and have yielded satisfactory
results when predicting exchange rates. Nwankwo (2014) forecasted the
exchange rate between the Naira and dollar, using Akaike's Information
Criterion (AIC) as a measure of performance. Diagnostic testing revealed
that the ARIMA(1,0,0) model was the best fit for the data.

Although the ARIMA model has the advantage of ease of implementation and
flexibility, it fails to capture the non-linearity and volatility
present in exchange rate data. Over time, the ARIMA model has evolved to
cater for a wider variety of data and to compensate for some of its
shortcomings. Some of the most popular versions of the ARIMA model that
has been implemented in exchange rate forecasting is the Seasonal ARIMA
(SARIMA) and Fractional ARIMA (ARFIMA) model.

The SARIMA model was introduced to capture the periodic behaviour of
data and extends the ARIMA class by including a term for seasonal
differencing. Etuk, Wokoma, and Moffat (2013) modelled the Naira/CFA
Franc exchange rate which exhibited monthly seasonality using an
additive SARIMA model, to demonstrate that it can be a useful fit for
exchange rate data which displays seasonality. Their results showed that
the SARIMA model adequately described the variation in the exchange rate
series.

The ARFIMA model generalises the ARIMA model in that the degree required
to make the data stationary can assume any real value, and is no longer
restricted to the integer domain. The ARFIMA model has the ability to
capture the dependence between observations that are widely spread apart
in time (Cheung 1993). This makes the model parsimonious since it can
capture long memory in data as well as short term dynamics (Cheung
1993). Cheung (1993) fitted the ARFIMA model to examine five exchange
rates and found that there was strong evidence of long memory in the
exchange rate time series.

Generalised autoregressive conditional heteroskedasticity (GARCH) models
were later developed due to the failure of ARIMA models to capture the
volatility in financial markets (Anastasakis and Mort 2009). Fahimifard
et al. (2009) applied a GARCH model to the Rial/USD and Rial/EUR to
compare the forecasts yielded by the GARCH and ARIMA model over varying
forecast horizons. Their results illustrated that the GARCH model
outperformed the ARIMA model when using the Root Mean Square Error
(RMSE), Mean Square Error (MSE) and Mean Absolute Deviation (MAD) as a
performance criteria.

Over time, financial forecasting methods have moved away from linear
models like ARIMA and GARCH, to soft computing techniques. These complex
techniques are non-linear and can fit complex time series more easily
(Castillo and Melin 2002). Unlike the ARIMA model, soft computing
techniques do not impose structural assumptions on the model apriori
(Castillo and Melin 2002). Some of the most commonly used artificial
intelligence methods used to forecast exchange rate data are Neural
Networks and Fuzzy Logistic Systems.

Artificial Neural Networks (ANNs) mimic the structure of the brain and
have had many successful applications in forecasting exchange rates.
They are data driven, can adapt to non-stationary environments and can
approximate any continuous function (Khashei and Bijari 2011). In a
study done by Fahimifard et al. (2009), the ANN was found as an
effective way to improve the forecasts of exchange rates. Superior
results were produced when compared to the ARIMA and GARCH model using
the RMSE, MSE and MAD as a measure of performance.

Fuzzy Logistic Systems (FLSs) were initially developed to solve problems
involving linguistic terms and have successfully been used in financial
forecasting (Khashei, Bijari, and Ardali 2009). Fuzzy logic tries to
imitate human reasoning and the decision-making process and allows for
finer rather than discrete decisions to be provided. Santos, Costa, and
Santos Coelho (2007) investigated how well FLSs and ANNs perform
compared to the traditional ARMA and GARCH model. He examined the
forecasts of Brazilian exchange rate returns by considering different
frequencies of the series and comparing their one-step ahead forecasts.
By analysing accuracy statistics, he found that FLSs and ANNs achieved
higher returns based on the forecasts they produced. Similar results
were found by Khashei, Bijari, and Ardali (2009) when he analysed the
predictive capabilities of FLSs, ANNs, the traditional ARIMA model and a
Fuzzy ARIMA model.

Although ANNs have been broadly applied in financial forecasting, the
process of building them is a complex task and there is no consistent
method of design compared to the traditional Box-Jenkins ARIMA model.
The performance of ANNs is sensitive to many modelling factors such as
the number of input nodes included and the size of the training sample
chosen (Zhang and Hu 1998). Like ANNs, there is no systematic approach
for designing FLSs and they are only understandable when simple.
Although FLSs has the advantage over ARIMA models in that they can be
applied to data with few observations available, it yields acceptable
rather than accurate results and are more suitable for problems which do
not require high accuracy.

The traditional ARIMA model produces less superior forecasts than its
hybrid forms and other complex non-linear techniques, however its
forecasts are still satisfactory. It is a simple model that is easy to
implement and has a consistent method of model design and selection.
ARIMA models are more robust and efficient than complex structural
models in relation to short-run forecasting. The fact that they have
been used as the foundation for more advanced models and have commonly
been used as a benchmark for comparison, justifies it as a good starting
point to compare it to Facebook's forecasting method, Prophet, that was
recently released.

The techniques that have been considered for exchange rate forecasting
thus far, require the analyst to have vocational knowledge about time
series. Prophet differs from traditional time series models in that it
is flexible and can be customised by a large number of non-experts who
have little knowledge about time series, however have domain knowledge
about the data generating process. Prophet allows for a large number of
forecasts to be produced across a variety of problems and consists of a
robust evaluation system that allows for a large number of forecasts be
evaluated and compared. This is Facebook's definition of forecasting at
scale.

Prophet's Bayesian approach to forecasting allows the analyst to
incorporate their expert knowledge into the model building process and
has produced significantly improved forecasts compared to the ARIMA
model. Taylor and Letham (2017) forecasted the number of events on
Facebook using Prophet. The time series was impacted by holidays, had
strong multi-period seasonality, and a piecewise trend. The forecasts
produced by Prophet were compared to common forecasting techniques such
as exponential smoothing, ARIMA, the seasonal naive, and the naive
model. While exponential smoothing and the seasonal naive model were
quite robust, the ARIMA forecasts were fragile. No model besides Prophet
accounted for the dips around holidays and the upward trend of the time
series towards later observations. If Prophet produces forecasts that
are as good as the ARIMA model when forecasting BTC/ZAR, it can be
compared to more complex forecasting methods that are less tractable
such as the hybrid models and machine learning techniques seen earlier.

The dataset used in this study consists of the daily closing prices of
BTC/ZAR over the period 24 January 2016 to 17 July 2017. This comprises
of a total of 541 trading days and was the chosen time period for
analysis due to constraints in obtaining data over a longer period. The
data was obtained from Bitcoincharts.

Bitcoin is of specific interest as it presents an interesting parallel
to traditional exchange rate markets. The cryptocurrency is built on a
decentralised system and as a result its value cannot be directly
influenced by a central authority (Fantazzini et al. 2016). In addition,
the Bitcoin market has attracted attention worldwide and is currently
the leading cryptocurrency, with awareness and adoption of the currency
growing over time (Fantazzini et al. 2016). Its novelty makes it a
highly volatile hence speculative market and provides an opportunity for
forecasting.

\section{\texorpdfstring{Forecasting Methods
\label{Forecasting Methods}}{Forecasting Methods }}\label{forecasting-methods}

\subsection{The ARIMA model}\label{the-arima-model}

The ARIMA model expresses the process \{\(y_t\)\} as a function of the
weighted average of past values of the process and lagged values of the
residuals. The weighted average of the past p values of the process
represents an autoregressive (AR) process of order p.~It feeds back past
values of the process into the current value, inducing correlation
between all lags of the process. The weighted average of the q lagged
residuals represents a moving average (MA) process of order q. The
purpose of mixing the MA process with the AR process is to reduce the
large number of past values required by AR processes and to control for
the autocorrelation which it creates between lagged values of the
process. The combination of the AR(p) and MA(q) process results in a
more parsimonious model, and forms a stationary autoregressive moving
average (ARMA(p,q)) process defined as:

@@

The ARIMA(p,d,q) model generalises the ARMA model in that it includes
both stationary and non-stationary processes. The parameter d is the
degree of differencing required to render the process stationary. If d
is equal to zero the process is stationary and equivalent to an ARMA
model, and if d is strictly positive the process requires differencing
to make it stationary. The ARIMA model can be defined succinctly using
the backward shift operator B, which shifts the process back by one unit
of time, and is defined as \(By_t = y_{t-1}\). The ARIMA model has the
form (Hyndman and Athanasopoulos 2014):

@@

\subsection{The Prophet model}\label{the-prophet-model}

Prophet is similar to a Generalized Additive Model (GAM) - an additive
regression model that consists of non-linear and linear regression
functions applied to predictor variables (Taylor and Letham 2017). The
decomposable model is of the form:

@@

Prophet, like the GAM, frames the forecasting problem as a curve fitting
exercise and uses backfitting to find the regression functions. This
allows for the model to be fitted quickly and missing values and large
outliers to be handled elegantly. The regression model also provides
model flexibility and allows the analyst to interactively change model
parameters (Taylor and Letham 2017). The growth component of Prophet may
be modelled as a linear or non-linear function of time. Linear growth is
modelled by a piecewise constant function while non-linear growth is
modelled similar to population growths which use a logistic growth model
(Taylor and Letham 2017). A time-varying upper limit may be specified
for logistic growth, at which point the forecasts will saturate. This
carrying capacity allows the analyst to incorporate their prior
knowledge about the maximum obtainable growth level such as the total
market or population size into the model. Prophet accounts for changes
in the trajectory of this trend by automatically detecting and selecting
changepoints in the data at which the growth rate is allowed to change.
These changepoints have a Laplace prior distribution placed on them and
its scale parameter may be used to adjust the flexibility of the trend
and to choose how aggressively the model should follow historical trend
changes. Analysts may also adjust the number of potential changepoints
included or manually specify their location. This allows non-experts
with knowledge about events that may affect the growth rate to use the
parameter as a knob to either increase or decrease the number of
changepoints included (Taylor and Letham 2017). It also allows for the
analyst to add changepoints which the automatic selection procedure may
have missed or remove changepoints when the model is overfitting
historical trends (Taylor and Letham 2017).

@@

@@

@@

In this paper, the out-of-sample forecasts of ARIMA and Prophet are
compared over varying forecast horizons. The data is first split into a
training set and test set. The training set starts on the 24th January
2016 and ends on the 16th January 2017, while the test set starts on the
17th January 2017 and ends on the 17th July 2017. The Box-Jenkins
methodology is then used to select the correct ARIMA model while
Prophet's semi-automatic procedure is used for model selection. The
selected ARIMA and Prophet model are then fitted to the training set and
rolling window forecasts are made 1 day, 30 days and 90 days ahead. This
allows us to examine the forecast horizon effect. The Model Confidence
Set and Mincer-Zarnowitz test is then used to evaluate the forecasts
produced by ARIMA and Prophet against the test set. Since there is no
consensus on which accuracy statistic best measures the performance of
forecasting techniques, the most popular criteria, the Root Mean Squared
Error (RMSE) is employed. The results obtained from the ARIMA model will
be used as a benchmark for comparison. This process is then repeated
based on data which has outliers and missing values present.

@@

Box, Jenkins, and Reinsel (1970) proposed a set of guidelines that can
be followed when selecting ARIMA models. It consists of a four-stage
iterative process in which:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The process is either transformed or differenced to de-trend and
  stabilise the variance of the data.
\item
  The autocorrelation and partial autocorrelation plots are used to
  determine the order of p and q.
\item
  The parameters of the model are then estimated.
\item
  A diagnostic check is performed to ensure that the residuals are a
  white noise process.
\end{enumerate}

If the residuals are not white noise, steps 2-4 are repeated until a
satisfactory model is identified. On the contrary, if the diagnostic
check reveals that the residuals are random, the developed model will be
the final model used for forecasting.
